---
title: "220420_MicroPouch_Cleaning_metadata"
author: "Sebastian Mølvang Dall"
date: "20/4/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Loading Data

```{r}
library(tidyverse)
library(janitor)
library(ggpubr)
library(lubridate)
library(haven)
library(labelled)
```



# Redcap metadata
## Patients
Loading metadata csv files - downloaded from REDcap

>> There are discrepencies between redcap and downloaded csv file!!


```{r}
metadata <- read_dta("../../../MicroPouch/data/metadata/stata/metadata_mp.dta") %>% 
  mutate(project = "MP") %>% 
  fill(randomization_group)

metadata_op <- read_dta("../../../MicroPouch/data/metadata/stata/metadata_op.dta") %>% 
  mutate(project = "OP")

metadata_np <- read_dta("../../../MicroPouch/data/metadata/stata/metadata_np.dta") %>% 
  mutate(project = "NP") %>% 
  rename(study_id = record_id)

```




### Joining the 3 datasets

All projects are concatenated into one variable. The projects are distinguished by the `project` variable.

```{r}
m_all <- bind_rows(metadata, metadata_op, metadata_np) %>% 
  arrange(study_id) %>% 
  relocate(project, .after = study_id)
```


### Renaming columns
* selecting relevant pouch data
* uniting "duplicate" pdai columns

the there are two versions of pdai questionnaire in the dataset. One from the inklusion and followup with pouch inspection, and one from the different treatment dates.


```{r}
m_sel <- m_all %>% 
  select(study_id, project, randomization_group, redcap_event_name,
         crp, leu, calp, stool_fre,
         contains("pdai"), starts_with("pouchdys"), starts_with("sibdq"),
         starts_with("diet"))
```


#### Calculating pdai score and add to m_p
* **The score is not calculated for "OP" since the pdai_e and h is not completed**

```{r}
m_pdai <- m_sel %>% 
  select(study_id, project, redcap_event_name, randomization_group, 
         starts_with("pdai"), -pdai_complete, -pdai_date) %>% 
  filter(redcap_event_name %in% c("inclusion_arm_1", "followup_30day_arm_1"),
         project == "NP") %>%
  group_by(study_id) %>% 
  replace(is.na(.), 0) %>% 
  mutate(pdai_score = rowSums(across(where(is.numeric)))) %>% 
  select(study_id, project, redcap_event_name, pdai_score) %>% 
  full_join(m_sel) %>% 
  arrange(study_id, project)
  
```



#### Deselecting irrelevant columns
* Some columns contains irrelevant information such as height and age in the diet questionnaire.

> deselect "timestamp", "complete", and "date column".

> deselect non-recoded diet questions


```{r}
m_deselect <- m_pdai %>% 
  select(-c(diet_date:diet_q1b),
         -c(diet_q3:diet_q7),
         -contains("complete"),
         -contains("date"),
         -contains("timestamp")
         )
```




### mutate and rename columns for joining later

This chunk prepares the dataframe to be concatenated with the lab-metadata later.
* Patient name is changed.
* redcap_event_name variables are renamed.
* randomization_group (placebo, FMT) is changed and renamed group.

This dataframe is going to be concatenated with the lab metadata.

```{r}
m_tmp <- m_deselect %>% 
  ungroup() %>% 
  # Changing study_id tags
  mutate(
    study_id = as.character(study_id),
    study_id = if_else(nchar(study_id) == 1,
                       paste0("pt00", study_id),
                       paste0("pt0", study_id)
                       ),
           
         # Renaming event names
         redcap_event_name = str_replace(redcap_event_name, "\\_arm\\_1", ""), 
         
  ) %>% 
  # Changing treatment variables in recap_event_name
  mutate(
    sub_label = stringr::str_split(redcap_event_name, "_"),
    tmp_label = paste0(map_if(.x = sub_label, .p = ~is.vector(.x), 2), "_", map_if(.x = sub_label, .p=~is.vector(.x), 1)),
    redcap_event_name = if_else(str_detect(tmp_label, "treatment"), tmp_label, redcap_event_name)
  ) %>%
  # Changing followup in redcap_event_name
  mutate(
    redcap_event_name = str_replace(redcap_event_name, "\\_days", "d"),
    redcap_event_name = str_replace(redcap_event_name, "day", "d"),
    redcap_event_name = str_replace(redcap_event_name, "\\_months", "m"),
    redcap_event_name = str_replace(redcap_event_name, "\\_month", "m"),
    redcap_event_name = str_replace(redcap_event_name, "\\_p", ""),
    redcap_event_name = str_replace(redcap_event_name, "follow\\_up", "followup")
  ) %>% 
  rename(
    id = study_id,
    stage = redcap_event_name
    
  ) %>% 
  mutate(randomization_group = if_else(randomization_group == 1,
                                      "placebo",
                                      "FMT"),
         randomization_group = replace_na(randomization_group, "FMT")
         ) %>% 
  rename(group = randomization_group)
```


## FMT-donors


```{r}
metadata_fmt <- read_dta("../data/metadata/stata/metadata_fmt.dta")
```


### Isolating relevant redcap event

Redcap_event_filter 1 is for the first article! In this project event 2 and 3 were used. Most of the metadata is inklusion analysis of pathological gut species. As patients are included these are not relevant and have been deselected.

```{r}
d_fmt_event_filter <- metadata_fmt %>% 
  mutate(redcap_repeat_instrument = if_else(redcap_repeat_instrument == "", "info", redcap_repeat_instrument),
         redcap_repeat_instance = if_else(redcap_repeat_instrument == "info", 2, redcap_repeat_instance)) %>% 
  filter(redcap_repeat_instance != 1) %>% 
  select(donor_id:gender,bmi_v2,
         starts_with("diet"))
```

### Renaming of donors and columns

```{r}
d_fmt_renaming <- d_fmt_event_filter %>% 
  mutate(donor_id = as.character(donor_id),
         donor_id = if_else(nchar(donor_id) == 1, paste0("do00", donor_id), paste0("do0", donor_id)),
         donor_id = if_else(str_detect(string = donor_id, pattern = "asym"), "asym004", donor_id)) %>% 
  rename(id = donor_id)
```

### Deleting columns that are NA across

```{r}
d_fmt_no_NA <- d_fmt_renaming %>% 
  filter(redcap_repeat_instrument %in% c("info", "diet_questionnaire")) %>% 
  group_by(id) %>% 
  filter(n_distinct(redcap_repeat_instrument) == 2)


d_fmt_no_NA
```

```{r}
unique(filter(d_full, project == "donor_screening")$id)
```



# Lab-relevant metadata
## Donor Batch Overview and Barcode
### Load Barcodes
```{r}
barcode_metadata <- read_csv("../data/metadata/csv/barcode_metadata.csv")
```



### Overview

A donor batch is associated with a date and a donor number
```{r}
DoBa_overview <- read_csv("../data/metadata/csv/DoBa_oversigt.csv")

DoBa_o_c <- DoBa_overview %>%
  mutate(Donor = gsub("MTKF ", "do", Donor),
         Donor = gsub(" ","_",Donor))

DoBa_o_c <- DoBa_overview %>%
  mutate(Donor = gsub("MTKF ", "", Donor),
         Donor = ifelse(nchar(Donor) > 3, "asym004", paste0("do",Donor)))


DoBa_o_l <- DoBa_o_c %>%
  pivot_longer(cols = !Donor, names_to = c("batch_donation_number"),
               values_to = "batch_name") %>%
  mutate(batch_donation_number = gsub(".donation", "", batch_donation_number),
         batch_donation_number = as.numeric(batch_donation_number)) %>%
  filter(!is.na(batch_name)) %>%
  rename(donor_number = Donor)

```




* Match donor names with `DoBa_o_c`
```{r}
DoBa <- barcode_metadata %>%
  select(donor_number, batch_donation_number, barcode_clean) %>%
  full_join(DoBa_o_l, by = c("donor_number", "batch_donation_number")) %>%
  filter(!is.na(batch_donation_number)) %>%
  mutate(batch_name = paste0("(",batch_name,")"))

#write_csv(DoBa, file = "../data/metadata/csv/DoBa_bo.csv")
```




## Donor and patient match overview

>**The last mutate corrects a date for do020 batch (17.09.19) corrected from (17.19.19)**


```{r}
PaDo_overview <- read_csv("../data/metadata/MicroPouch patient donor batch oversigt april2022.csv")

PaDo_c <- PaDo_overview %>%
  mutate(MicroPouch_NP = ifelse(grepl("(np)", Patient),T, F),
         MicroPouch_open = ifelse(grepl("(o)", Patient),T, F),
         ) %>%
  mutate(Patient = gsub(pattern = "[[:punct:]]np[[:punct:]]",replacement = "",x = Patient),
         Patient = gsub(pattern = "[[:punct:]]o[[:punct:]]",replacement = "",x = Patient),
         Patient = gsub(pattern = " ", replacement = "", x = Patient))

PaDo_l <- PaDo_c %>%
  pivot_longer(Donor_1_batch_1:Donor_4_batch_2, names_to = "donor_batch", values_to = "batch") %>%
  filter(!is.na(batch)) %>% 
  separate_rows(batch, sep = " ") %>%
  mutate(cat = ifelse(nchar(batch) == 10, "batch", "donor")) %>%
  pivot_wider(names_from = cat, values_from = batch) %>%
  mutate(donor = ifelse(donor != "asymp004", paste0("do",donor), "asym004"),
         study = ifelse(MicroPouch_NP, "NP",
                        ifelse(MicroPouch_open, "OP", "MP"))) %>%
  select(-c(MicroPouch_NP, MicroPouch_open)) %>%
  mutate(batch = ifelse(test = grepl("do020", donor), yes = ifelse(test = grepl("(17.19.19)", batch), yes = "(17.09.19)", no = batch), no = batch)) # Error in patient data overview.csv

#write_csv(PaDo_c,"../data/metadata/csv/PatientDonor_match.csv")

```


## Join information in DoBa and PaDo
This will create a dataframe for donor batch connection. 

```{r}
d_pado <- DoBa %>%
  rename(batch = batch_name, donor = donor_number) %>%
  full_join(PaDo_l, by = c("donor", "batch")) %>% 
  mutate(stage = str_sub(donor_batch, 1,7),
         stage = str_replace(stage, "Donor", "treatment")) %>% 
  mutate(treatment_batch_number = str_sub(donor_batch,9,-1)) %>% 
  mutate(stage = str_replace(stage, "1", "5"),
         stage = str_replace(stage, "2", "10"),
         stage = str_replace(stage, "3", "15"),
         stage = str_replace(stage, "4", "21")
         )
```




# Extraction koncentration and completion of study.

I need to figure out who handed in a fecal sample. Those are the persons with sample_dates in the overview metadata. This has to be done for PaPr, PaOP and PaNP
Thereafter I need to join these tables, and connect them with the EXT_conc file. Lastly I need to make a plot showing each fecal sample and its concentration. This should preferably also be done for donors and screenings.

## PaPr overview

> I have changed value pt001, kontrol30dg, 05.11.1 -> 05.11.19

```{r}
PaPr_overview <- read_csv("../data/metadata/csv/PaPr_overview.csv")


PaPr_c <- PaPr_overview %>%
  mutate(patient = gsub("[a-zA-z]", "", patient),
         patient = gsub("[ -]","",patient),
         patient = paste0("pt",patient)) %>%
  #relocate(Kontrol_30dg, .after = `4_behandling`) %>% 
  clean_names() %>% 
  rename(inclusion = inklusion, 
         treatment_5 = x1_behandling, 
         treatment_10 = x2_behandling,
         treatment_15 = x3_behandling,
         treatment_21 = x4_behandling,
         followup_30d = kontrol_30dg,
         followup_1m = kontrol_1md,
         followup_3m = kontrol_3md,
         followup_6m = kontrol_6md,
         followup_12m = kontrol_12md)

PaPr_l <- PaPr_c %>%
  pivot_longer(-patient, names_to = "stage", values_to = "patient_batch") %>%
  mutate(patient_batch = ifelse(!is.na(patient_batch), paste0("(",patient_batch,")"),patient_batch)) %>%
  mutate(project = "MP") %>% 
  #filter(!is.na(patient_batch)) %>% 
  group_by(patient) %>% 
  mutate(fecal_donation_number = row_number())
```




## PaOP overview

* load
* clean up patient column
* relocate kontrol_30dg
* make long format and change patient_batch.
```{r}
PaOP_overview <- read_csv("../data/metadata/csv/PaOP_overview.csv")


PaOP_c <- PaOP_overview %>%
  mutate(patient = gsub("[a-zA-z]", "", patient),
         patient = gsub("[ -]","",patient),
         patient = paste0("pt",patient)) %>%
  #relocate(Kontrol_30dg, .after = `4_behandling`) %>% 
  rename(kontrol_3md = `Kontrol 3mdr`) %>% 
  clean_names() %>% 
  rename(inclusion = inklusion, 
         treatment_5 = x1_behandling, 
         treatment_10 = x2_behandling,
         treatment_15 = x3_behandling,
         treatment_21 = x4_behandling,
         followup_30d = kontrol_30dg,
         followup_1m = kontrol_1md,
         followup_3m = kontrol_3md)

PaOP_l <- PaOP_c %>%
  pivot_longer(-patient, names_to = "stage", values_to = "patient_batch") %>%
  mutate(patient_batch = ifelse(!is.na(patient_batch), paste0("(",patient_batch,")"),patient_batch)) %>% 
  mutate(project = "OP") %>% 
  group_by(patient) %>% 
  mutate(fecal_donation_number = row_number())
```



## PaNP

* load
* clean up patient column
* make long format and change patient_batch.

```{r}
PaNP_overview <- read_csv("../data/metadata/csv/PaNP_overview.csv")


PaNP_c <- PaNP_overview %>%
  mutate(patient = gsub("[a-zA-z]", "", patient),
         patient = gsub("[ -]","",patient),
         patient = paste0("pt",patient)) %>% 
  clean_names() %>% 
  rename(inclusion = inklusion, 
         treatment_5 = x1_behandling, 
         treatment_10 = x2_behandling,
         treatment_15 = x3_behandling,
         treatment_21 = x4_behandling,
         followup_30d = kontrol_30dg,
         followup_1m = kontrol_1md,
         followup_3m = kontrol_3md,
         followup_6m = kontrol_6md,
         followup_12m = kontrol_12md)

PaNP_l <- PaNP_c %>%
  pivot_longer(-patient, names_to = "stage", values_to = "patient_batch") %>%
  mutate(patient_batch = ifelse(!is.na(patient_batch), paste0("(",patient_batch,")"),patient_batch)) %>% 
  mutate(project = "NP") %>% 
  group_by(patient) %>% 
  mutate(fecal_donation_number = row_number())
```


## Joining the overview tables

Here I join the tables and filter out any patient_batch that was not received.

Change stage names to english

Add the fecal_batch_donation for later left_join

```{r}
stage_fct <- c("inclusion", 
         "treatment_5", 
         "treatment_10",
         "treatment_15",
         "treatment_21",
         "followup_30d",
         "followup_1m",
         "followup_3m",
         "followup_6m",
         "followup_12m",
         "drop_out")

d_overview <- bind_rows(PaPr_l, PaOP_l, PaNP_l) %>% 
  filter(!is.na(patient_batch)) %>% 
  mutate(stage = factor(stage, levels = stage_fct, labels = stage_fct))
```



## Adding library concentrations

ext_conc contains the extraction conc for each sample.

```{r}
ext_conc <- read_csv("../data/metagenomes_lab/220531_metagenome_libraries.csv")

d_ext <- full_join(d_overview, ext_conc, by = c("patient", "fecal_donation_number", "project")) %>% 
  mutate(patient = as.factor(patient),
         patient = fct_rev(patient),
         project = factor(project, levels = c("MP","OP","NP")),
         amount = ext_conc * 80,
         ext_conc = round(ext_conc, 2))

```


#### Fill in donor batch information

```{r}

d_fill_donorbatch <- d_pado %>% 
  select(donor, barcode_clean, batch_donation_number, batch) %>% 
  distinct() %>% 
  rename(patient = donor,
         sample_barcode = barcode_clean,
         fecal_donation_number=batch_donation_number,
         donor_batch = batch) %>% 
  right_join(d_ext, by = c("sample_barcode", "patient","fecal_donation_number")) %>% 
  unite("batch", c(patient_batch, donor_batch), na.rm = T)

```



## Adding treatment_1-4(5-21)_donor_batch
```{r}
d_treat <- d_pado %>% 
  select(-c(donor_batch,batch, barcode_clean)) %>% 
  pivot_wider(names_from = treatment_batch_number, values_from = batch_donation_number) %>% 
  rename(patient = Patient,
         project = study)
```



### gathering all metadata and deselecting irrelevant columns

Also some tidying before joining.

* Assinging project to donor screening
* Uniting date and fecal_batch_date -> redundant
* Tidying negative and positive controls
* changing the LibID column because of error from excel sheet

```{r}
d_full <- d_fill_donorbatch %>% 
  rename(fecal_batch_date = batch) %>% 
  left_join(d_treat, by = c("patient", "stage", "project")) %>% 
  rename(id = patient) %>%
  # putting donor screening and Asym into are project
  mutate(project = if_else(str_detect(sample_barcode, "DoBa"), true = "donor_batch",
                           false = if_else(str_detect(sample_barcode, "DoSc"), true = "donor_screening", false = project))) %>% 
  mutate(project = if_else(str_detect(sample_barcode, "Asym"), true = "donor_batch", false = project)) %>% 
  # uniting date and fecal_batch_date
  mutate(date = if_else(!is.na(date), paste0("(",date,")"), NA_character_)) %>% 
  unite('fecal_batch_date', c(fecal_batch_date, date), remove = T, na.rm = T, sep = "") %>% 
  mutate(fecal_batch_date =  if_else(fecal_batch_date == "", NA_character_, fecal_batch_date)) %>% 
  # Giving all rows a variable called is_sample = TRUE for easy isolation later
  mutate(is_sample = TRUE) %>% 
  # Tidying negative and postive controls
  mutate(sample_barcode = if_else(paste0(new_row,new_column) == "E12", "NEG_5", sample_barcode),
         sample_barcode = if_else(paste0(new_row,new_column) == "F12", "PCR_POS", sample_barcode),
         sample_barcode = if_else(is.na(id),
                                  true = paste0(library_plate, "_", sample_barcode),
                                  false = sample_barcode)) %>% 
  # Excel sheet error correction LIB00007: MP014 -> MP013 
  mutate(LibID = if_else(library_plate == "LIB00007", str_replace(LibID, "MP014", "MP013"), LibID)) %>% 
  
  # Deselecting irrelevant columns
  select(-c(scan_time, barcode, orientation_barcode,))
```

## Correcting stages
Apparently follow up is continued even though the a drop_out has occurred. This interferes with the other projects. I will try to solve this:

> A problem occurs because same patient in different projects have 2 fecal donations with the same date. However can they be in two projects at one time???. pt9 (MP+OP) MP_followup_6m == OP_followup_1m. PaPr00000082MP = PaOp00000052MP?

```{r}

open_patients <- d_full %>% 
  filter(project == "OP") %>% 
  select(id, project) %>% 
  distinct(id)


d_stage_correction <- d_full %>% 
  # Convert fecat_batch_date to lubridate formaet
  mutate(fecal_batch_date = str_replace(fecal_batch_date, "\\(",""),
         fecal_batch_date = str_replace(fecal_batch_date, "\\)",""),
         fecal_batch_date = dmy(fecal_batch_date)) %>% 
  # Move follow-ups after drop out to OP instead
  select(id, project, stage, fecal_batch_date) %>% 
  filter(project %in% c("MP", "NP", "OP")) %>% 
  pivot_wider(names_from = stage, values_from = fecal_batch_date) %>% 
  pivot_longer(-c(id, project, drop_out), names_to = "stage", values_to = "fecal_batch_date") %>% 
  filter(!is.na(fecal_batch_date)) %>% 
  mutate(project = if_else(
    condition = project == "MP" & drop_out < fecal_batch_date & id %in% open_patients$id, 
    true = "OP",
    false = project
  )) %>% 
  pivot_wider(names_from = stage, values_from = fecal_batch_date) %>% 
  pivot_longer(drop_out:followup_3m, names_to = "stage", values_to = "fecal_batch_date")
  
```




```{r}
#write_csv(d_full, "../data/metadata/csv/sample_metadata.csv")
```

## Full metadata

```{r}
m_full_metadata <- full_join(d_full,m_tmp, by = c("id", "stage", "project")) %>% 
  mutate(is_sample = if_else(is.na(is_sample), FALSE, is_sample)) %>% 
  relocate(is_sample, .after = project)


#write_delim(m_full_metadata, "../data/metadata/csv/full_metadata.csv", delim = ";")
```


