---
title: "Alpha/Beta diversity"
author: "Sebastian Mølvang Dall"
date: '2022-08-05'
output: html_document
---

## libraries
```{r}
library(tidyverse)
library(vegan)
library(ggpubr)
library(mplibrary)

source(file = "R/functions/metadata_functions.R")
# source(file = "R/functions/bioinformatics.R")
source(file = "R/functions/gg_functions.R")
```

## metadata + mikrobiom data
```{r}
metadata <- load_metadata()

metaphlan <- load_metaphlan("species")

project_filter <- "NP"
```


### Isolate metadata in metaphlan table
First step is isolation of samples in colnames(metaphlan) in metadata sheet. Thereafter, relevant columns are selected.

```{r}
metadata_in_metaphlan <- metadata %>%
  filter(LibID %in% colnames(metaphlan))

t_metaphlan <- transposeMetaphlan(metaphlan)
```
```{r}
metadata_NP <- isolateProjectMetadata(metadata, project_filter = project_filter)
metadata_NP_with_data <- isolateProjectMetadata(metadata_in_metaphlan, project_filter = project_filter)

remission_relapse <- metadata_NP_with_data %>%
  filter(!is.na(pdai_score)) %>%
  group_by(id) %>%
  filter(n_distinct(stage) == 2) %>%
  select(id, stage, group, pdai_score) %>%
  pivot_wider(names_from = stage, values_from = pdai_score) %>%
  mutate(remission = if_else(inclusion - followup_30d > 2 & followup_30d <= 7, "remission", "relapse")) %>%
  select(id, group, remission)

metadata_NP_remission <- left_join(metadata_NP_with_data, remission_relapse)
```

### Calculating species richness and shannon diversity

Species richness is defined as species with a relative abundance >0. Alpha diversity was calculated using Shannon Index.

```{r}
# Species richness
species_richness <- metaphlan %>%
  calculateSpeciesRichness()

# Shannon diversity
shannon_diversity <- tibble(LibID = rownames(t_metaphlan), shannon = diversity(as.matrix(t_metaphlan), index = "shannon"))

alpha_diversity_df <- full_join(species_richness, shannon_diversity)

metadata_with_alpha <- metadata_in_metaphlan %>%
  left_join(alpha_diversity_df)

metadata_NP_remission_alpha <- left_join(metadata_NP_remission, alpha_diversity_df)
```


## GGplot Alpha
### FMT/PLACEBO

Depending on the `project_filter` tag two different ggplots are created. The ggplot code can be found in `functions/gg_functions.R`.

```{r}
gg_alpha_rich <- gg_alpha(metadata_NP_remission_alpha, alpha_metric = "richness") +
  labs(y = "Number of Species") +
  ylim(0, 100) +
  theme(legend.position = "none")

gg_alpha_shan <- gg_alpha(metadata_NP_remission_alpha, alpha_metric = "shannon") +
  ylim(0, 4) +
  labs(y = "Shannon Diversity")

ggarrange(gg_alpha_rich, gg_alpha_shan, ncol = 2, labels = c("A", "B"), common.legend = T, legend = "bottom")

# ggsave(paste0("../figures/",project_filter,"_alpha_rich_shan.png"), device = "png", width = 8)
```

## Beta-Diversity


### BRAY CURTIS
Distance calculated as the median bray curtis distance from patient to set of donor samples received.

```{r}
source("functions/beta_diversity_function.R")
```


```{r}

if (project_filter == "NP") {
  stages <- grep("Donor", stages_all, invert = T, value = T)
} else {
  stages <- c("Pre", "Post")
}

metadata_beta_diversity_bray <- beta_diversity(metadata_MP, t_metaphlan, transform = T, stages = stages, seed = 1)
```

```{r}
metadata_beta_bray <- createMetadataWBetadiversity(metadata_NP, t_metaphlan, method = "bray", transform = TRUE)

metadata_beta_bray_compared <- compareFMT2ActualDonor(projectmetadata = metadata_NP, metadata_beta_bray)

metadata_beta_bray_summarised <- summariseBetaDiversityOutput(metadata_beta_bray_compared)
```

```{r}
gg_beta_bray <- gg_beta(metadata_beta_bray_summarised, "bray")
```



### Sorensen

```{r}
metadata_beta_sor_summarised <- createMetadataWBetadiversity(metadata_NP, t_metaphlan, "sorensen", FALSE) %>%
  createANDcombineComparedOutputs(projectmetadata = metadata_NP) %>%
  summariseBetaDiversityOutput()

gg_beta_sor <- gg_beta(metadata_beta_sor_summarised, "sorensen") +
  theme(legend.position = "right")
```

```{r}
ggarrange(gg_beta_sor, gg_beta_bray, ncol = 2, labels = c("A", "B"), common.legend = T, legend = "bottom")
# ggsave(paste0("../figures/",project_filter,"_beta_diversity.png"), device = "png", height = 5, width = 8)
```
