---
title: "heatmap"
author: "Sebastian Mølvang Dall"
date: '2022-08-10'
output: html_document
---

## Loading libraries

```{r}
library(tidyverse)
library(ggpubr)
source("functions/metadata_functions.R")
```

## Loading Data

**Important!:The MicroPouchNP heatmap is in the bottom. Follow the headers**

```{r}
metadata <- load_metadata()

metaphlan <- load_metaphlan(taxonomic_level = "genus")

project_filter = "NP"
```

```{r}
metadata_all <- prepare_metadata(metadata)

metadata_in_metaphlan <- prepare_metadata(metadata) %>% 
  filter(LibID %in% colnames(metaphlan))


metaphlan_long <- metaphlan %>% 
  select(-NCBI_tax_id) %>% 
  pivot_longer(-clade_name, names_to = "LibID") %>% 
  arrange(LibID)
```

### Filtering for relevant patients and donors

Just as in alpha_beta_diversity.rmd, the donors are filtered such that only donors used for the available patients are used.

```{r}
if (project_filter == "MP") {
  stages = stages_pre_post


  metadata_relevant <- IsolateProjectAndDonor(metadata_in_metaphlan, project_filter = "MP") %>% 
    mutate(sample_group = paste0(group, "_", x_axis),
           sample_group = if_else(str_detect(sample_group, "FMT_Donor"), str_remove(sample_group, "FMT_"), sample_group)) %>% 
    filter(x_axis %in% stages) %>% 
    group_by(id) %>% 
    filter(n_distinct(x_axis) == 2 | x_axis == "Donor")
  
  n_genera = 25
  
} else if (project_filter== "NP") {
  stages = stages_all

  metadata_relevant <- IsolateProjectAndDonor(metadata_in_metaphlan, project_filter = "NP") %>% 
    mutate(sample_group = paste0(group, "_", x_axis),
           sample_group = if_else(str_detect(sample_group, "FMT_Donor"), str_remove(sample_group, "FMT_"), sample_group)) %>% 
    filter(x_axis %in% stages) 
  
  n_genera = 35
}


```

#### Adding taxonomy to metadata table

```{r}
metadata_relevant_with_tax <- left_join(metadata_relevant, metaphlan_long)
```

## heatmap visualization

### Donor, Pre, heatmap

#### Top genera
This chunk finds the most abundant genera by calculating the weighted mean of each taxonomic clade in each group, `Pre FMT`, `Pre Placebo`, and `Donor`. Each group is weighted by the inverse number of samples in each group. Since there are the least amount of samples in the `Donor` group their relative abundance will have a higher weight in the mean. The purpose is to find engraftment from the donor microbiome to the patients therefore the `Donor` microbiome gets a higher weight. In the end the top genera will be arranged according to their prevalence in the `Donor` group. 


```{r}
metadata_weigth <- metadata_relevant_with_tax %>% 
  mutate(sample_group = paste0(x_axis, " ", group),
         sample_group = if_else(str_detect(sample_group, "Donor FMT"), str_remove(sample_group, " FMT"), sample_group)) %>% 
  group_by(donor_batch, sample_group) %>% 
  mutate(hellinger = sqrt(value/sum(value)),
         sum = sum(value)) %>% 
  ungroup()


top_genera_weighted <- metadata_weigth %>%
  filter(x_axis %in% c("Pre", "Donor")) %>% 
  group_by(id, x_axis, clade_name) %>% 
  summarise(hellinger = mean(hellinger)) %>% 
  group_by(x_axis, clade_name) %>% 
  summarise(
    weight = n(),
    mean_hellinger = mean(hellinger)
  ) %>% 
  group_by(clade_name) %>% 
  summarise(weighted_mean_hellinger = weighted.mean(mean_hellinger, weight)) %>% 
  arrange(desc(weighted_mean_hellinger)) %>% 
  head(n_genera)


top_genera_weighted_donor <- metadata_weigth %>%
  filter(clade_name %in% top_genera_weighted$clade_name, x_axis == "Donor") %>% 
  group_by(id, clade_name) %>% 
  summarise(hellinger = mean(hellinger)) %>% 
  group_by(clade_name) %>% 
  summarise(mean_hellinger = mean(hellinger)) %>% 
  arrange(desc(mean_hellinger))

```



#### hclust

For better visualization purposes the samples pre treatment are clustered based on their microbiome. Here a distance matrix using Bray-Curtis is calculated on Hellinger transformed relative abundances and clustered using the "ward D.2" algorithm, which tries to minimize within cluster variance using sum of squared(D2) errors.


```{r}
t_metaphlan <- metaphlan %>% 
  select(-NCBI_tax_id) %>% 
  pivot_longer(-clade_name, names_to = "LibID") %>% 
  pivot_wider(names_from = clade_name, values_from = value) %>% 
  filter(LibID %in% metadata_weigth$LibID) %>% 
  arrange(LibID) %>% 
  column_to_rownames(var = "LibID") %>% 
  as.data.frame()



t_metaphlan_id <- metadata_weigth %>% 
  filter(str_detect(sample_group, "Pre") | sample_group == "Donor") %>% 
  arrange(LibID) %>% 
  select(id, LibID, group, sample_group, x_axis) %>% 
  distinct(LibID, .keep_all = T) %>% 
  left_join(
    rownames_to_column(t_metaphlan,"LibID")
  )
  
t_metaphlan_long_id_donor_avg <- t_metaphlan_id %>% 
  pivot_longer(!id:x_axis, names_to = "clade_name", values_to = "relative_abundance") %>% 
  group_by(id, x_axis, clade_name) %>% 
  summarise(relative_abundance = mean(relative_abundance))

t_metaphlan_wide_id_donor_avg <- t_metaphlan_long_id_donor_avg %>% 
  ungroup() %>%
  select(!x_axis) %>%
  pivot_wider(names_from = clade_name, values_from = relative_abundance) %>% 
  column_to_rownames("id")

bray_curtis_dist <- vegan::vegdist(vegan::decostand(t_metaphlan_wide_id_donor_avg, method = "hellinger"))

hclust_ward <- hclust(bray_curtis_dist, method = "ward.D2")

ward_dendogram <- as.dendrogram(hclust_ward)
ward_order <- order.dendrogram(ward_dendogram)

plot(ward_dendogram)


id_cluster_order <- t_metaphlan_long_id_donor_avg %>% 
  left_join(metadata_weigth) %>% 
  mutate(
    id = factor(
      id,
      levels = hclust_ward$labels[ward_order]
    )
  ) %>% 
  group_by(id, clade_name, sample_group) %>% 
  distinct(id, .keep_all = T)
  
```

## MP heatmaps
### Donor / Pre heatmap

Pre and donor heatmap clustered with hclust ward.D2 method. donor batches have been grouped by average relative abundance.

```{r}

gg_heat_inc <- id_cluster_order %>% 
  filter(clade_name %in% top_genera_weighted$clade_name) %>% 
  mutate(clade_name = factor(clade_name, levels = top_genera_weighted_donor$clade_name),
         x_axis = factor(x_axis, levels = c("Pre", "Donor")),
         sample_group = factor(sample_group, levels = c("Pre FMT", "Pre placebo", "Donor")),
         relative_abundance = if_else(relative_abundance < 10, round(relative_abundance, 1), floor(relative_abundance))) %>% 
  ggplot(aes(x = id, y = fct_rev(clade_name),  fill = hellinger, label = relative_abundance)) +
    geom_tile(color = "white") +
    scale_fill_gradient(low = "white", high = "red") +
    geom_text(color = "black") +
    facet_grid(. ~ sample_group, scales = "free_x", space = "free") +
    labs(x = "", y= "") +
    plot_theme +
    theme(axis.text.x = element_text(angle = 90, vjust = -0.0001),
          legend.position = "none")
gg_heat_inc

ggsave("../figures/heatmap_pre_donor_weight-hellinger.png", device = "png", width = 12, height = 10)
```



### Post heatmap


```{r}

metadata_tax_post <- metadata_weigth %>% 
  filter(x_axis == "Post") %>% 
  arrange(LibID)



t_metaphlan_post_id <- metadata_tax_post %>%  
  select(id, LibID, group, sample_group,x_axis) %>% 
  distinct(LibID, .keep_all = T) %>% 
  left_join(
    rownames_to_column(t_metaphlan,"LibID")
  ) %>% 
  select(!LibID: x_axis) %>% 
  column_to_rownames("id")
  
```

```{r}
post_id_cluster_order <- metadata_tax_post %>% 
  mutate(
    id = factor(
      id,
      levels = str_subset(hclust_ward$labels[ward_order], "do", negate = T)
    )
  )


post_id_cluster_order %>% 
  filter(clade_name %in% top_genera_weighted$clade_name) %>% 
  mutate(clade_name = factor(clade_name, levels = top_genera_weighted_donor$clade_name),
         sample_group = factor(sample_group, levels = c("Post FMT", "Post placebo")),
         value = if_else(value < 10, round(value, 1), floor(value))) %>% 
  ggplot(aes(x = id, y = fct_rev(clade_name),  fill = hellinger, label = value)) +
    geom_tile(color = "white") +
    scale_fill_gradient(low = "white", high = "red") +
    geom_text(color = "black") +
    facet_grid(. ~ sample_group, scales = "free_x", space = "free") +
    labs(x = "", y= "") +
    plot_theme +
    theme(axis.text.x = element_text(angle = 90, vjust = -0.0001))


ggsave("../figures/heatmap_post_weight-hellinger.png", device = "png", width = 9, height = 10)

```


This aligns Pre and post heatmap for visualisation.
```{r}
gg_heat_post <- post_id_cluster_order %>% 
  filter(clade_name %in% top_genera_weighted$clade_name) %>% 
  mutate(clade_name = factor(clade_name, levels = top_genera_weighted_donor$clade_name),
         sample_group = factor(sample_group, levels = c("Post FMT", "Post placebo")),
         value = if_else(value < 10, round(value, 1), floor(value))) %>% 
  ggplot(aes(x = id, y = fct_rev(clade_name),  fill = hellinger, label = value)) +
    geom_tile(color = "white") +
    scale_fill_gradient(low = "white", high = "red") +
    geom_text(color = "black") +
    facet_grid(. ~ sample_group, scales = "free_x", space = "free") +
    labs(x = "", y= "") +
    plot_theme +
    theme(axis.text.x = element_text(angle = 90, vjust = -0.0001),
          axis.text.y = element_blank(),
          axis.ticks.y = element_blank())
# 
# ggarrange(gg_heat_inc, gg_heat_post, ncol = 2, widths = c(3,2))
# 
# ggsave("../figures/heatmap_pre_post_hellinger.png", device = "png", width = 21, height = 10)
```

### Group by treatment

```{r}
id_cluster_order <- t_metaphlan_long_id_donor_avg %>% 
  right_join(metadata_weigth) %>% 
  mutate(
    id = factor(
      id,
      levels = hclust_ward$labels[ward_order]
    )
  ) %>% 
  group_by(id, clade_name, sample_group, x_axis, group) %>% 
  summarise(value = mean(value),
            hellinger = mean(hellinger)) %>% 
  filter(clade_name %in% top_genera_weighted$clade_name) %>% 
  mutate(clade_name = factor(clade_name, levels = top_genera_weighted_donor$clade_name),
         x_axis = factor(x_axis, levels = c("Pre", "Post", "Donor")),
         sample_group = factor(sample_group, levels = c("Pre FMT","Post FMT", "Pre placebo", "Post placebo", "Donor")),
         relative_abundance = if_else(value < 10, round(value, 1), floor(value))) 



gg_heat_fmt <- id_cluster_order %>% 
  filter(group == "FMT") %>% 
  ggplot(aes(x = id, y = fct_rev(clade_name),  fill = hellinger, label = relative_abundance)) +
    geom_tile(color = "white") +
    scale_fill_gradient(low = "white", high = "red") +
    geom_text(color = "black") +
    facet_grid(. ~ sample_group, scales = "free_x", space = "free") +
    labs(x = "", y= "") +
    plot_theme +
    theme(axis.text.x = element_text(angle = 90, vjust = -0.0001),
          legend.position = "none")
gg_heat_fmt

ggsave("../figures/heatmap_fmt_donor.png", device = "png", width = 12, height = 10)
```


```{r}
gg_heat_placebo <- id_cluster_order %>% 
  filter(group == "placebo") %>% 
  ggplot(aes(x = id, y = fct_rev(clade_name),  fill = hellinger, label = relative_abundance)) +
    geom_tile(color = "white") +
    scale_fill_gradient(low = "white", high = "red") +
    geom_text(color = "black") +
    facet_grid(. ~ sample_group, scales = "free_x", space = "free") +
    labs(x = "", y= "") +
    plot_theme +
    theme(axis.text.x = element_text(angle = 90, vjust = -0.0001),
          legend.position = "none")

gg_heat_placebo
ggsave("../figures/heatmap_placebo.png", device = "png", width = 9, height = 10)
```



## NP heatmap 


```{r}
labels <- c("Pre", paste0("Day ", c(5,10,16,28)), "Post", paste0("Follow-up ", c(1,3,6, 12), "M"), paste0("Donor_b", 1:12))
stages_withdonorbatch <- c(grep("Donor", stages, invert = T, value = T), paste0("Donor_b", 1:12))

metadata_weigth %>% 
  filter(clade_name %in% top_genera_weighted$clade_name) %>% 
  # group_by(id, x_axis, clade_name) %>% 
  # summarise(hellinger = mean(hellinger),
  #           value = mean(value)) %>% 
  mutate(
     clade_name = factor(clade_name, levels = top_genera_weighted_donor$clade_name),
     x_axis = if_else(x_axis == "Donor", paste0(x_axis, "_b", fecal_donation_number), x_axis),
     x_axis = factor(x_axis, levels = stages_withdonorbatch, labels = labels ),
     value = if_else(value < 10, round(value, 1), floor(value)),
     id = factor(id, levels = c( "pt001", "pt002", "pt006", "asym004"))
     ) %>%
  ggplot(aes(x = x_axis, y = fct_rev(clade_name),  fill = hellinger, label = value)) +
    geom_tile(color = "white") +
    scale_fill_gradient(low = "white", high = "red") +
    geom_text(color = "black") +
    facet_grid(. ~ id, scales = "free_x", space = "free") +
    labs(x = "", y= "") +
    plot_theme +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.35, hjust = 1),
          axis.text = element_text(size = 12))

ggsave("../figures/NP_heatmap_hellinger_alldonorbatches.png", device = "png", width = 20, height = 10)
```

### Treatment schema

```{r}
donorTreatment <- metadata_relevant %>% 
  pivot_longer(batch_1:batch_3, names_to = "batch", values_to = "batch_number") %>% 
  filter(!is.na(batch_number)) %>% 
  select(id, x_axis, donor, batch, batch_number) %>% 
  mutate(batch = str_replace(batch, "_", " "),
         x_axis = factor(x_axis, levels = stages_withdonorbatch, labels = labels))


donorTreatment %>% 
  ggplot(aes(x = batch, y = id, label = batch_number, group = id)) +
  geom_line() +
  geom_point(size = 10, color = "lightblue") +
  geom_text() +
  facet_grid(.~x_axis, scales = "free_x") +
  plot_theme +
  labs(y = "") +
  theme(panel.grid.major.x = element_blank())

ggsave("../figures/NP_donorbatch_treatment.png", device = "png")
```




