---
title: "Ordination plot"
author: "Sebastian Mølvang Dall"
date: '2022-08-24'
output: html_document
---


## libraries and functions

```{r}
library(tidyverse)
library(ampvis2)
library(RColorBrewer)
source("functions/metadata_functions.R")
source("functions/bioinformatics.R")
```


## Load data

```{r}
metadata <- load_metadata()

metaphlan <- load_metaphlan("fulltax_species")

project_filter = "MP"
```



### Prepare metadata

```{r}
metadata_in_metaplan <- prepare_metadata(metadata) %>% 
  filter(LibID %in% colnames(metaphlan))

metadata_MP <- prepare_metadata(metadata) %>%
  filter(LibID %in% colnames(metaphlan)) %>%
  IsolateProjectAndDonor(project_filter = "MP") %>% 
  relocate(LibID) %>% 
  mutate(sample_group = paste0(x_axis, " ",group),
         sample_group = if_else(str_detect(sample_group, "Donor FMT"), str_remove(sample_group, " FMT"), sample_group))
```


### Prepare tax data

```{r}
metaphlan_otu <- metaphlan %>%
 select(-NCBI_tax_id) %>%
 select(clade_name, metadata_MP$LibID) %>% 
 mutate(OTU = paste0("OTU",row_number())) %>% 
 relocate(OTU)


otutable <- metaphlan_otu %>% 
  select(-clade_name)


taxtable <- metaphlan_otu %>% 
  select(OTU, clade_name) %>% 
  separate(clade_name, sep = "\\|", into = c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species")) %>% 
  relocate(OTU, .after = "Species")
```


## Load ampvis


```{r}
d <- amp_load(
  otutable = otutable,
  metadata = metadata_MP,
  taxonomy = taxtable
)
```


### Filter ampvis for pre post

```{r}
stages = stages_pre_post

metadata_pre_post <- metadata_MP %>% 
  filter(x_axis %in% stages, x_axis != "Donor") %>% 
  group_by(id) %>%
  filter(n_distinct(x_axis) == 2) %>% 
  ungroup()
```

```{r}
d_pre_post <- d %>% 
  amp_subset_samples(LibID %in% metadata_pre_post$LibID)
```

#MP
### Ordination

```{r}
pca <- d %>% 
  amp_subset_samples(LibID %in% metadata_pre_post$LibID | x_axis == "Donor") %>% 
  amp_ordinate(
  type = "pca",
  transform = "hellinger",
  sample_color_by = "sample_group",
  sample_shape_by = "x_axis",
  species_plot = F,
  species_nlabels = 5,
  species_label_taxonomy = "Genus",
  filter_species = 0,
  sample_plotly = "id",
  sample_colorframe = F
) + 
  geom_line(aes(group = id), color = "grey70") +
  theme(legend.title = element_blank()) +
  scale_color_brewer(palette = "Paired")

pt014_do016 <- pca$data %>% 
  filter(id == "pt014" & x_axis == "Post" | donor_batch == "do016_4")
pca +
  geom_text(data = pt014_do016, aes(x = PC1, y = PC2+0.04, label = id), color = "red", size = 3)

#ggsave("../figures/amp_pca_hellinger_all.png", device = "png", width = 7, height = 5)
```



```{r}
paired_2_5 = brewer.pal(n = 5, "Paired")[2:5]

rda_ord_pt <- d_pre_post %>% 
  #amp_filter_samples(x_axis == "Post") %>% 
  amp_ordinate(
  
  type = "rda",
  constrain = "sample_group",
  transform = "hellinger",
  sample_color_by = "sample_group",
  sample_colorframe = T,
  sample_colorframe_label = "sample_group",
  species_plot = F,
  envfit_show = T,
  detailed_output = T
)

rda_ord_pt$plot +
  labs(fill = "", color = "") +
  scale_color_manual(values = paired_2_5) +
  scale_fill_manual(values = paired_2_5) +
  theme(legend.position = "none")


#ggsave("../figures/amp_rda_hellinger_no_donor.png", device = "png")
```





```{r}

metadata_all_stages <- metadata_MP %>% 
  filter(project %in% c("MP", "donor_batch"), x_axis %in% stages_all, group != "placebo") %>% 
  mutate(x_axis = factor(x_axis, levels = stages_all)) %>% 
  arrange(id, fecal_donation_number)

d_MP_all <- amp_load(
  otutable = otutable,
  metadata = metadata_all_stages,
  taxonomy = taxtable
)



d_MP_all %>% 
  amp_ordinate(
  type = "pca",
  transform = "hellinger",
  sample_color_by = "id",
  sample_shape_by = "project",
  #sample_plotly = "donor_batch",
  species_plot = F,
  species_nlabels = 5,
  species_label_taxonomy = "Genus",
  filter_species = 0,
  sample_colorframe = F
) +
  geom_path(aes(group = id), color = "grey70")



  
```



Next chunck calculates the difference between centroids for groups in the PCA.
`envfit()` calculates the regression of supplementary variables on the ordination axes of unconstrained ordination. Furthermore `envfit()` can test the significance of this test by permutation (that is shuffling of the supplementary variable).

For factors significance of the fit is calculated by testing if the variation within centroids are significantly different from total spread of the data. That is $$r^{2} = 1 - \frac{ss_w}{ss_t} $$, where $$ss_{w}$$ and $$ss_t$$ are within-group and total sums of squares, respectively. To test if this fit is significant `envfit()` and `pairwise.factorfit()` uses permutations of the data to test if this explanation of variance is significant.

*It is crucial* that test for significance using `envfit()` is done on unconstrained ordination. If the test is done on constrained data the ordination axis already contain the information of the variable and as a consequence a randomly assigned variable can be highly significant (because the axis already contain the information).



```{r}
t_metaphlan <- metaphlan %>% 
  dplyr::select(!NCBI_tax_id) %>% 
  pivot_longer(-clade_name, names_to = "LibID") %>% 
  pivot_wider(names_from = clade_name, values_from = value) %>% 
  filter(LibID %in% metadata_pre_post$LibID) %>% 
  arrange(LibID) %>% 
  column_to_rownames("LibID") %>% 
  abundance_filter(abundance_threshold = 10^-34) %>% 
  vegan::decostand("hellinger") %>% 
  rownames_to_column("LibID")


MetadataMetaphlan <- metadata_pre_post %>% 
  dplyr::select(LibID) %>% 
  ungroup() %>% 
  left_join(t_metaphlan) %>% 
  column_to_rownames("LibID") %>% 
  as.data.frame()

constrain <- metadata_pre_post %>%
  arrange(LibID) %>% 
  select(sample_group) 
  

RDA <- rda(MetadataMetaphlan ~ sample_group, data = constrain)
envfit(RDA, env = constrain)

pca <- rda(MetadataMetaphlan)
envfit(pca ~ constrain$sample_group)

pairwise.factorfit(pca, constrain$sample_group)

```

#NP
## NP and MP donor PCA

```{r}
metadata_NP <- prepare_metadata(metadata) %>%
  filter(LibID %in% colnames(metaphlan)) %>%
  IsolateProjectAndDonor(project_filter = "NP") %>% 
  relocate(LibID) %>% 
  mutate(sample_group = paste0(x_axis, " ",group),
         sample_group = if_else(str_detect(sample_group, "Donor FMT"), str_remove(sample_group, " FMT"), sample_group),
         project = "NP")

metadata_MP <- prepare_metadata(metadata) %>%
  filter(LibID %in% colnames(metaphlan)) %>%
  IsolateProjectAndDonor(project_filter = "MP") %>% 
  relocate(LibID) %>% 
  mutate(sample_group = paste0(x_axis, " ",group),
         sample_group = if_else(str_detect(sample_group, "Donor FMT"), str_remove(sample_group, " FMT"), sample_group),
         project = "MP")

metadata_NP_MP <- bind_rows(metadata_MP, metadata_NP) %>% 
  mutate(sample_group = paste0(sample_group, " ", project))

```

```{r}
metaphlan_otu <- metaphlan %>%
 select(-NCBI_tax_id) %>%
 select(clade_name, metadata_NP_MP$LibID) %>% 
 mutate(OTU = paste0("OTU",row_number())) %>% 
 relocate(OTU)


otutable <- metaphlan_otu %>% 
  select(-clade_name)


taxtable <- metaphlan_otu %>% 
  select(OTU, clade_name) %>% 
  separate(clade_name, sep = "\\|", into = c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species")) %>% 
  relocate(OTU, .after = "Species")


d <- amp_load(
  otutable = otutable,
  metadata = metadata_NP_MP,
  taxonomy = taxtable
)
```


```{r}
stages = stages_pre_post

metadata_NP_MP_stages <- metadata_NP_MP %>% 
  filter(x_axis %in% stages)

d_subset <- d %>% 
  amp_subset_samples(LibID %in% metadata_NP$LibID | x_axis == "Donor", x_axis %in% stages, id != "pt002")
```



```{r}
d_subset %>% 
  amp_ordinate(
  type = "pca",
  transform = "hellinger",
  sample_color_by = "sample_group",
  species_plot = T,
  species_nlabels = 7,
  species_label_taxonomy = "Genus",
  sample_shape_by = "x_axis"
) + 
  geom_line(aes(group = id), color = "grey70") +
  theme(legend.title = element_blank()) +
  scale_color_brewer(palette = "Paired")

ggsave("../figures/NP_PCA.png", device = "png", width = 6, height = 6)


```

```{r}
d_subset %>% 
  amp_heatmap(tax_show = 35,
              normalise = F,
              tax_aggregate = "Genus")
```


## MP + Asym

```{r}
metadata <- prepare_metadata(metadata) %>%
  filter(LibID %in% colnames(metaphlan)) %>% 
  relocate(LibID)


metaphlan_otu <- metaphlan %>%
 select(-NCBI_tax_id) %>%
 select(clade_name, metadata$LibID) %>% 
 mutate(OTU = paste0("OTU",row_number())) %>% 
 relocate(OTU)

otutable <- metaphlan_otu %>% 
  select(-clade_name)

d <- amp_load(
  otutable = otutable,
  metadata = metadata,
  taxonomy = taxtable
)

```


```{r}
metadata_sub <- metadata %>% 
  filter(LibID %in% metadata_MP$LibID | id == "asym004") %>% 
  filter( stage %in% c(NA, "inclusion", "followup_30d"), project != "donor_screening", group %in% c(NA, "FMT")) %>% 
  group_by(id) %>% 
  filter(n() > 1)
```





```{r}
pca <- d %>% 
  amp_subset_samples(LibID %in% metadata_sub$LibID) %>% 
  amp_ordinate(
  type = "pca",
  transform = "hellinger",
  sample_color_by = "project",
  sample_shape_by = "stage",
  species_plot = F,
  species_nlabels = 5,
  species_label_taxonomy = "Genus",
  filter_species = 0,
  sample_colorframe = F
) + 
  geom_line(aes(group = id), color = "grey70") +
  theme(legend.title = element_blank()) +
  scale_color_brewer(palette = "Paired")

pt014_do016 <- pca$data %>% 
  filter(id == "pt014" & x_axis == "Post" | donor_batch == "do016_4")
pca +
  geom_text(data = pt014_do016, aes(x = PC1, y = PC2+0.04, label = id), color = "red", size = 3)

ggsave("../figures/amp_pca_hellinger_fmt_donor.png", device = "png", width = 7, height = 5)
```


